# Descriptive Statistics

Descriptive statistics are a set of summary measures that provide a concise overview of a dataset. They help us understand the characteristics and properties of the data without delving into complex statistical analyses. Some commonly used descriptive statistics include the mean, standard deviation, and median.

To illustrate the concept, let's consider fish length measurement $x$. Each fish is identified by a subscript, and their lengths are denoted as follows:

$$
x_1 = 15.9\\
x_2 = 15.1\\
x_3 = 21.9\\
x_4 = 13.3\\
x_5 = 24.4\\
$$Often times, we use subscript $i$ (or any character you like) instead of actual number to indicate a given data point. For example, we write fish length $x_i$ for individual $i$. Alternatively, instead of writing each individual data point, we can represent them as a vector using boldface, denoted as $\pmb{x}$. In this case, the vector $\pmb{x}$ can be expressed as:

$$
\pmb{x} = \{15.9, 15.1, 21.9, 13.3, 24.4\}
$$

Now let's see how we represent summary statistics of vector $\pmb{x}$.

## Central Tendency

### Measures of Central Tendency

Central tendency measures (Table \@ref(tab:central)) provide insights into the typical or central value of a dataset. There are three commonly used measures:

-   **Arithmetic Mean**. This is the most commonly used measure of central tendency. It represents the *additive* average of the data. To calculate the arithmetic mean, you sum up all the values and divide the total by the number of data points. It can be heavily influenced by extreme values, known as outliers.

-   **Geometric Mean**. The geometric mean is a *multiplicative* average. It is always smaller than the arithmetic mean and less sensitive to unusually large values. However, it is not applicable when the data contain negative values.

-   **Median**. The median is the value that separates the higher half from the lower half of the dataset. It represents the 50th percentile data point. The median is less affected by outliers compared to the arithmetic mean. To calculate the median, you arrange the data in ascending order and select the middle value if the dataset has an odd number of values. If the dataset has an even number of values, you take the average of the two middle values.

| Measure                   | Equation                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                  |
|------------------|-------------------------------------------------------------|
| Arithmetic mean $\mu$     | $$                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                     \frac{\sum_i^N x_i}{N}                                                                                                                                                             
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                     $$                                                                                                                                                                                 |
| Geometric mean $\mu_{ge}$ | $$                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                     (\prod_i^N x_i)^{\frac{1}{N}}                                                                                                                                                      
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                     $$                                                                                                                                                                                 |
| Median $\mu_{med}$        | $$                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                     \begin{aligned}                                                                                                                                                                    
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                     &x_{(\frac{N + 1}{2}^\text{th})} &&\text{if N is odd number}\\                                                                                                                     
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                     &\frac{1}{2}[x_{(\frac{N}{2}^\text{th})} + x_{(\frac{N}{2} + 1^\text{th})}] &&\text{if N is even number}                                                                           
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                     \end{aligned}$$                                                                                                                                                                    |

: (#tab:central) Common measures of central tendency. $N$ refers to the number of data points.

### R Exercise

To learn more about these measures, let's create vectors $\pmb{x} = \{15.9, 15.1, 21.9, 13.3, 24.4\}$ and $\pmb{y} = \{15.9, 15.1, 21.9, 53.3, 24.4\}$ -- $\pmb{y}$ is identical to $\pmb{x}$ but contains one outlier value. How does this make difference? To construct vectors in R, we use `c()`, a function that stands for "construct." Below is the script:

```{r construct}

# construct vectors x and y
x <- c(15.9, 15.1, 21.9, 13.3, 24.4)
y <- c(15.9, 15.1, 21.9, 53.3, 24.4)
```

Confirm you constructed them correctly:

```{r construct-confirm}
x
y
```

Cool! Now we can calculate summary statistics of `x` and `y`.

**Arithmetic mean**

While R has a function for arithmetic `mean()`, let's try to calculate the value from scratch:

```{r arithmetic-mean}

# for vector x
n_x <- length(x) # the number of elements in x = the number of data points
sum_x <- sum(x) # summation for x
mu_x <- sum_x / n_x # arithmetic mean
print(mu_x) # print calculated value

# for vector y; we can calculate directly too
mu_y <- sum(y) / length(y)
print(mu_y) # print calculated value
```

Compare with outputs from `mean()` :

```{r mean-buildin}

print(mean(x))
print(mean(y))
```

**Geometric Mean**

Unfortunately, there is no build-in function for geometric mean $\mu_{ge}$ in R (as far as I know; there are packages though). But, we can calculate the value from scratch again:

```{r geometric-mean}

# for vector x
prod_x <- prod(x) # product of vector x; x1 * x2 * x3...
n_x <- length(x)
mug_x <- prod_x^(1 / n_x) # ^ means power
print(mug_x)

# for vector y
mug_y <- prod(y)^(1 / length(y))
print(mug_y)
```

**Median**

Lastly, let's do the same for median:

```{r median}

# for vector x
x <- sort(x) # sort x from small to large
index <- (length(x) + 1) / 2 # (N + 1)/2 th index as length(x) is an odd number
med_x <- x[index]
print(med_x)

# for vector y
y <- sort(y) # sort y from small to large
med_y <- y[(length(y) + 1) / 2]
print(med_y)
```

Compare with outputs from `median()`

```{r median-builtin}
print(median(x))
print(median(y))
```

## Variation

### Measures of Variation

Variation measures (Table \@ref(tab:variation)) provide information about the spread of data points.

-   **Variance.** Variance is a statistical measure that quantifies the spread or dispersion of a dataset. It provides a numerical value that indicates how far individual data points in a dataset deviate from the mean or average value. In other words, variance measures the average squared difference between each data point and the mean. **Standard deviation (SD) is the square root of variance**.

-   **Inter-Quantile Range.** The interquartile range (IQR) is a statistical measure that provides information about the spread or dispersion of a dataset, specifically focusing on the middle 50% of the data. It is a robust measure of variability that is less affected by outliers compared to variance.

-   **Median Absolute Deviation.** Median Absolute Deviation (MAD) is similar to variance, but provides a robust estimation of variability that is less affected by outliers compared to variance. MAD is defined as the median of the absolute deviations from the data's median.

-   **Coefficient of Variation.** The coefficient of variation (CV) is a statistical measure that expresses the relative variability of a dataset in relation to its mean. It is particularly useful when comparing the variability of different datasets that have different scales or units of measurement.

-   **MAD/Median.** MAD/Median is a statistical measure used to assess the relative variability of a dataset without assuming any specific distribution or parametric model. CV is sensitive to outliers because of its reliance on the arithmetic mean. However, MAD/Median is robust to this issue.

| Measure                                                                                        | Equation                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                          |
|---------------------------|------------------------------------------------------------|
| Variance $\sigma^2$ (standard deviation $\sigma$)                                              | $$                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        \frac{\sum_i^N (x_i - \mu)^2}{N}                                                             
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        $$                                                                                           |
| Inter-Quantile Range IQR ($x_l$ and $x_h$ are $l^{th}$ and $h^{th}$ percentiles, respectively) | $$                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        |x_{l} - x_{h}|                                                                              
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        $$                                                                                           |
| Median Absolute Deviation (MAD)                                                                | $$                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        \text{Median}(|x_i-\mu_{med}|)                                                               
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        $$                                                                                           |
| Coefficient of Variation (CV)                                                                  | $$                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        \frac{\sigma}{\mu}                                                                           
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        $$                                                                                           |
| MAD/Median                                                                                     | $$                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        \frac{\text{MAD}}{\mu_{med}}                                                                 
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        $$                                                                                           |

: (#tab:variation) Common measures of variation. $N$ refers to the number of data points.

### R Exercise

**Variance, SD, and CV**

Let's try variance, SD, and CV:

```{r variance}
# for x
sqd_x <- (x - mean(x))^2 # sqared deviance
sum_sqd_x <- sum(sqd_x)

var_x <- sum_sqd_x / length(x)
sd_x <- sqrt(var_x) # sqrt(): square root
cv_x <- sd_x / mean(x)

print(var_x)
print(sd_x)
print(cv_x)

# for y
var_y <- sum((y - mean(y))^2) / length(y)
sd_y <- sqrt(var_y)
cv_y <- sd_y / mean(y)

print(var_y)
print(sd_y)
print(cv_y)
```

**IQR, MAD, and MAD/Median**

Let's try IQR, MAD, and MAD/Median. IQR can be defined for given percentiles. Here, let me use 25 and 75 percentiles as $x_l$ and $x_h$.

```{r iqr}

# for x
## IQR
x_l <- quantile(x, 0.25) # quantile(): return quantile values, 25 percentile
x_h <- quantile(x, 0.75) # quantile(): return quantile values, 75 percentile
iqr_x <- abs(x_l - x_h) # abs(): absolute value

## MAD
ad_x <- abs(x - mean(x))
mad_x <- median(ad_x)

## MAD/median
mad_m_x <- mad_x / median(x)

print(iqr_x)
print(mad_x)
print(mad_m_x)

# for y
## IQR
y_q <- quantile(y, c(0.25, 0.75)) # return as a vector
iqr_y <- abs(y_q[1] - y_q[2]) # y_q[1] = 25 percentile; y_q[2] = 75 percentile

## MAD
mad_y <- median(abs(y - mean(y)))

## MAD/median
mad_m_y <- mad_y / median(y)

print(iqr_y)
print(mad_y)
print(mad_m_y)
```

## Laboratory

### Comparing Central Tendency Measures

What are the differences of the three measures of central tendency? To investigate this further, let's perform the following exercise.

1.  Create a new vector `z` with length $1000$ as `exp(rnorm(n = 1000, mean = 0, sd = 0.1))`, and calculate the arithmetic mean, geometric mean, and median.

2.  Draw a histogram of `z` using functions `tibble()`, `ggplot()`, and `geom_histogram()`.

3.  Draw vertical lines of arithmetic mean, geometric mean, and median on the histogram with different colors using a function `geom_vline()` .

4.  Compare the values of the central tendency measures.

5.  Create a new vector `z_rev` as `-z + max(z) + 0.1`, and repeat step 1 -- 4.

### Comparing Variation Measures

Why do we have absolute (variance, SD, MAD, IQR) and relative measures (CV, MAD/Median) of variation? To understand this, suppose we have 100 measurements of fish weight in unit "gram." (`w` in the following script)

```{r gram}
w <- rnorm(100, mean = 10, sd = 1)
head(w) # show first 10 elements in w
```

Using this data, perform the following exercise:

1.  Convert the unit of `w` to "milligram" and create a new vector `m`.

2.  Calculate SD and MAD for `w` and `m`.

3.  Calculate CV and MAD/Median for `w` and `m`.
