```{r setup-02, include=FALSE}

pacman::p_load(tidyverse,
               foreach)
```

# Why statistics?

## The rationale

*Why do I need statistics in ecology in the first place?* - this is the first question I got when I stepped into the field of ecology. I thought this would be an easy question that someone could give me the answer right away, but actually it is a deep one. The short answer is "*we need statistics because we only have **partial** information about what we want to know.*" The long answer? See below.

## Example

```{r fish-data}
lake_size <- 15
set.seed(10)

## data set
df_plot <- tibble(x0 = runif(10, 0, lake_size - 1),
                  x1 = x0 + 1,
                  y0 = runif(10, 0, lake_size - 1),
                  y1 = y0 + 1)

df_fish <- tibble(x = runif(500, 0, lake_size),
                  y = runif(500, 0, lake_size))


df_fish_sub <- foreach(i = seq_len(nrow(df_plot)),
                       .combine = bind_rows) %do% {
  df_fish %>% 
    filter(x > df_plot$x0[i],
           x < df_plot$x1[i],
           y > df_plot$y0[i],
           y < df_plot$y1[i]) %>% 
    mutate(plot = i,
           count = nrow(.))
}
```

Let say we are interested in fish density in a lake (total area $15 \times 15 = 225~\text{m}^2$). We try to quantify it by catching fish in a certain area (1 m^2^ for example). Since there should be some variation in fish density among sampled points, we decided to repeat this `r nrow(df_plot)` times to take an average. Figure \@ref(fig:fish-plot) shows a graphical representation of this example.

```{r fish-plot, fig.cap="Lake fish density example. Points are fish, and white boxes are sampling sites with 1 sq-m"}

g_lake <- ggplot(data = df_fish_sub,
                 aes(x = x,
                     y = y)) +
  geom_rect(aes(xmin = 0,
                xmax = lake_size,
                ymin = 0,
                ymax = lake_size),
            fill = "steelblue") +
  geom_point(size = 0.5) +
  geom_rect(data = df_plot,
            aes(x = x0,
                y = y0,
                xmin = x0,
                xmax = x1,
                ymin = y0,
                ymax = y1),
            color = "white",
            fill = NA) +
  scale_x_continuous(limits = c(0, lake_size)) +
  scale_y_continuous(limits = c(0, lake_size)) +
  theme_minimal()

g_lake
```

Through this sampling, we got the following dataset:

```{r, echo = FALSE}

df_count <- df_fish_sub %>% 
  right_join(tibble(plot = 1:nrow(df_plot)),
             by = "plot") %>% 
  mutate(count = replace_na(count, 0)) %>% 
  distinct(plot, count) %>% 
  arrange(plot)

df_count
```

Based on this dataset, we may conclude the average fish density in this lake is `r round(mean(df_count$count), 2)`. However, how confident are you on this conclusion? The lake area is $15 \times 15 = 225~\text{m}^2$ but we sampled only $1~\text{m}^2 \times 10 = 10~\text{m}^2$! Indeed, we missed a lot of fish in the lake (Figure \@ref(fig:fish-plot-2)).

```{r fish-plot-2, fig.cap="True fish distribution in the lake"}
g_lake + 
  geom_point(data = df_fish,
             size = 0.5,
             color = grey(0, 0.2))
```

In ecological research (and almost all research field), we are rarely able to obtain the perfect dataset, i.e., no uncertainty in the data.
