# Structural Equation Modeling

In path analysis and structural equation modeling (SEM), we are interested in understanding the relationships among multiple variables simultaneously. Unlike simple regression, which models a single response as a function of one or more predictors, SEM allows researchers to specify networks of mediating pathways, providing a framework to quantify both direct and indirect contributions.

Learning Objectives:

-   Understand the conceptual difference between univariate regression, path analysis, and SEM.

-   Specify and fit path models using ecological data, identifying direct and indirect effects.

-   Interpret model outputs, including path coefficients, residual variances, and fit statistics.

-   Quantify and test the significance of hypothesized causal relationships in multivariate ecological systems.

```{r}
pacman::p_load(tidyverse,
               GGally,
               vegan,
               lavaan)
```

## Path analysis

```{r}
url <- "https://raw.githubusercontent.com/aterui/biostats/master/data_raw/data_foodweb.csv"
df_fw <- read_csv(url)
```

```{r}
df_fw %>% 
  select(-plot_id) %>% 
  ggpairs(
    progress = FALSE,
  )
```

```{r}
df_mite <- df_mite %>% 
  mutate(
    across(.cols = where(is.numeric),
           .fns = function(x) c(scale(x)))
  )

m <- '
  watr_cont ~ shrub
  log_rich ~ watr_cont + shrub
'

(fit <- sem(model = m,
           data = df_mite))
```

## Structural Equation Modeling

### Latent variable
