# Data Structure {.unnumbered}

## Overview {.unnumbered}

R has 6 basic **data types**.

-   character: `"aquatic"`, `"ecology"` (no order)
-   factor: similar to character, but has *levels* (alphabetically ordered by default)
-   numeric: `20.0` , `15.5`
-   integer: `3`, `7`
-   logical: `TRUE` , `FALSE`
-   complex: `1+2i` (complex numbers with real and imaginary parts)

These elements form one of the following **data structures**.

-   **vector**: a series of elements. A single data type is allowed in a single vector
-   **matrix**: elements organized into rows and columns. A single data type is allowed in a single matrix
-   **data frame**: looks similar to a matrix, but allows different data types in different columns

## Vector {.unnumbered}

### Create Vector {.unnumbered}

Below are examples of atomic character vectors, numeric vectors, integer vectors, etc. There are many ways to create vector data. The following examples use `c()`, `:`, `seq()`, `rep()`:

```{r, echo=TRUE}

#ex.1a manually create a vector using c()
x <- c(1,3,4,8)
x

#ex.1b character
x <- c("a", "b", "c")
x

#ex.1c logical
x <- c(TRUE, FALSE, FALSE)
x

#ex.2 sequence of numbers
x <- 1:5
x

#ex.3a replicate same numbers or characters
x <- rep(2, 5) # replicate 2 five times
x

#ex.3b replicate same numbers or characters
x <- rep("a", 5) # replicate "a" five times
x

#ex.4a use seq() function
x <- seq(1, 5, by = 1)
x

#ex.4b use seq() function
x <- seq(1, 5, by = 0.1)
x

```

### Check Features {.unnumbered}

R provides many functions to examine features of vectors and other objects, for example:

-   `class()` - return high-level data structure of the object
-   `typeof()` - return low-level data structure of the object
-   `attributes()` - metadata of the object
-   `length()` - number of elements in the object
-   `sum()` - sum of object's elements
-   `mean()` - mean of object's elements

**Numeric Vector**

```{r, echo=TRUE}

x <- c(1.2, 3.1, 4.0, 8.2)
x
class(x)
typeof(x)
length(x)
sum(x)
mean(x)

```

**Character Vector**

```{r}

y <- c("a", "b", "c")
class(y)
length(y)

```

### Access {.unnumbered}

**Element ID**\
Use brackets `[]` when accessing specific elements in an object. For example, if you want to access element #2 in the vector `x`, you may specify as `x[2]`:

```{r}

x <- c(2,2,3,2,5)
x[2] # access element #2
x[c(2,4)] # access elements #2 and 4
x[2:4] # access elements #2-4

```

**Equation**\
R provides many ways to access elements that suffice specific conditions. You can use mathematical symbols to specify what you need, for example:

-   `==` equal
-   `>` larger than
-   `>=` equal & larger than
-   `<` smaller than
-   `<=` equal & smaller than
-   `which()` a function that returns element \# that suffices the specified condition

The following examples return a logical vector indicating whether each element in x suffices the specified condition:

```{r, echo=TRUE}

# creating a vector
x <- c(2,2,3,2,5)

# ex.1a equal
x == 2

# ex.1b larger than
x > 2 

```

You can access elements that suffice the specified condition using brackets, for example:

```{r, echo=TRUE}

# ex.2a equal
x[x == 2]
# ex.2b larger than
x[x > 2]

```

Using `which()`, you can see which elements (i.e., \#) matches what you need:

```{r, echo=TRUE}

# ex.3a equal
which(x == 2) # returns which elements are equal to 2
# ex.3b larger than
which(x > 2)

```

## Matrix {.unnumbered}

### Create Matrix {.unnumbered}

Matrix is a set of elements (*single data type*) that are organized into rows and columns:

```{r, echo=TRUE}

#ex.1 cbind: combine objects by column
x <- cbind(c(1,2,3), c(4,5,6))
x

#ex.2 rbind: combine objects by row
x <- rbind(c(1,2,3), c(4,5,6))
x

#ex.3 matrix: specify elements and the number of rows (nrow) and columns (ncol)
x <- matrix(1:9, nrow = 3, ncol = 3)
x

```

### Check Features {.unnumbered}

R provides many functions to examine features of matrix data, for example:

-   `dim()` number of rows and columns
-   `rowSums()` row sums
-   `colSums()` column sums

**Integer Matrix**

```{r, echo=TRUE}

x <- matrix(1:9, nrow = 3, ncol = 3)
x
class(x)
typeof(x)
dim(x)

```

**Character Matrix**

```{r, echo=TRUE}

y <- matrix(c("a","b", "c", "d", "e", "f"), nrow = 3, ncol = 2)
y
class(y)
typeof(y)
dim(y)

```

### Access {.unnumbered}

When accessing matrix elements, you need to pick row(s) and/or column(s), for example:

```{r, echo=TRUE}

x <- matrix(1:9, nrow = 3, ncol = 3)
x
x[2,3] # access an element in row #2 and colum #3
x[2,] # access elements in row #2
x[c(2,3),] # access elements in rows #2 and 3
x[,c(2,3)] # access elements in columns #2 and 3

```

You can assess each element with mathematical expressions just like vectors:

```{r, echo=TRUE}

x == 2 # equal
x > 2 # larger than

```

However, care must be taken when accessing elements, as it will be automatically converted to vector data:

```{r}

x[x == 2] # equal
x[x > 2] # larger than

```

`which()` needs an additional argument to return both row and column #:

```{r}

which(x == 2, arr.ind = TRUE)
which(x > 2, arr.ind = TRUE)

```

## Data Frame {.unnumbered}

A data frame is a collection of elements organized into rows and columns, but it differs from a matrix in several ways.

-   It allows for the inclusion of *multiple data types* in different columns.
-   Each column in a data frame has a *name* associated with it.
-   You can access columns in a data frame by their respective names using the `$` operator.

The data frame is the most commonly used data structure when manipulating ecological data. When loading a dataset from a spreadsheet (which we will discuss later), it is automatically recognized as a data frame. Let's consider an example:

**Creating a data frame**\
In the following example, the variables `x` and `y` are organized into a single data frame named `df0`. The variables are renamed as part of the process of creating the data frame.

```{r, echo=TRUE}
# Create data frame
x <- c("Pristine", "Pristine", "Disturbed", "Disturbed", "Pristine") # Lake type
y <- c(1.2, 2.2, 10.9, 50.0, 3.0) # TSS: total suspended solids (mg/L)
df0 <- data.frame(LakeType = x, TSS = y) # x is named as "LakeType" while y is named as "TSS"
df0
```

**Call column names**

```{r, echo=TRUE}
colnames(df0) # call column names
```

**Access by columns**

```{r, echo=TRUE}
df0$LakeType # access LakeType
df0$TSS # access TSS
```

You can access elements like a matrix as well:

```{r, echo=TRUE}

df0[,1] # access column #1
df0[1,] # access row #1
df0[c(2,4),] # access row #2 and 4

```

## Exercise {.unnumbered}

### Vector {.unnumbered}

a.  Create three numeric vectors with length 3, 6 and 20, respectively. Vectors must be created using at least two different functions in R.
b.  Create three character vectors with length 3, 6 and 20, respectively. Vectors must be created using at least two different functions in R.
c.  Copy the following script to your R script and perform the following analysis:

-   Identify element IDs of `x` that are greater than `2.0`
-   Identify element values of `x` that are greater than `2.0`

```{r, eval = FALSE}
set.seed(1)
x <- rnorm(100)
```

### Matrix {.unnumbered}

a.  Create a numeric matrix with 4 rows and 4 columns. Each column must contain identical elements.
b.  Create a numeric matrix with 4 rows and 4 columns. Each row must contain identical elements.
c.  Create a character matrix with 4 rows and 4 columns. Each column must contain identical elements.
d.  Create a character matrix with 4 rows and 4 columns. Each row must contain identical elements.
e.  Copy the following script to your R script and perform the following analysis:

-   Identify element IDs of `x` that are greater than `2.0` (**specify row and column IDs**)
-   Identify element values of `x` that are greater than `2.0` and calculate the mean.

```{r, eval = FALSE}
set.seed(1)
x <- matrix(rnorm(100), nrow = 10, ncol = 10)
```

### Data Frame {.unnumbered}

a.  Create a data frame of 3 variables with 10 elements (name variables as `x`, `y` and `z`. `x` must be `character` while `y` and `z` must be `numeric`.
b.  Check the data structure (higher-level) of `x`, `y` and `z`
c.  Copy the following script to your R script and perform the following analysis:

-   Calculate the means of `temperature` and `abundance` for states `VA` and `NC` separately.

```{r, eval = FALSE}
set.seed(1)
x <- rnorm(100, mean = 10, sd = 3)
y <- rpois(100, lambda = 10)
z <- rep(c("VA", "NC"), 50)
df0 <- data.frame(temperature = x, abundance = y, state = z)
```
