```{r, include=F}
source(here::here("code/library.R"))
```

# Tidyverse {.unnumbered}

## Overview {.unnumbered}

R packages are collections of functions that extend the functionality of the R programming language. They provide a convenient way for users to access and utilize specialized tools for data analysis, visualization, statistical modeling, and more. Among them, `tidyverse` provides very useful functions for data manipulation and visualization.

`tidyverse::` bundles R packages designed to make data manipulation, exploration, and visualization more efficient and intuitive. Developed by Hadley Wickham and other contributors, the Tidyverse packages share a common philosophy and syntax, emphasizing a consistent and tidy data format. The core packages, such as `dplyr::`, `tidyr::`, and `ggplot2::`, provide powerful tools for data wrangling, reshaping, and creating visualizations.

In this Chapter, we will use `iris` data, which is available by default in R. While the default data structure `iris` is prepared in `data.frame()`, I convert this data format to `tibble()` in this exercise.

```{r}
library(tidyverse)
iris <- as_tibble(iris)
```

I cover the basics for those who are not familiar with `tidyverse::`, but there is plenty of documentation available for the tidyverse package.

-   [Tidyverse website](https://www.tidyverse.org/)
-   [dplyr](https://dplyr.tidyverse.org/)
-   [tidyr](https://tidyr.tidyverse.org/)
-   [ggplot2](https://r-graph-gallery.com/ggplot2-package.html)
-   [from Data to Viz](https://www.data-to-viz.com/)

## Data Manipulation {.unnumbered}

### Data Format {.unnumbered}

`dplyr::` and `tidyr::` (and others) offer useful functions for data manipulation. Key manipulations include: row/column manipulations, group operation, reshape, and more.

### Row Manipulation {.unnumbered}

`filter()`: select/remove rows

```{r, eval=F}
# single match "=="
filter(iris, Species == "virginica")

# multiple match "%in%"
filter(iris, Species %in% c("virginica", "versicolor"))

# except "!="
filter(iris, Species != "virginica")

# except multiple "!(x %in% c("a", "b"))
filter(iris, !(Species %in% c("virginica", "versicolor")))

# greater than ">"
filter(iris, Sepal.Length > 5)

# equal & greater than ">="
filter(iris, Sepal.Length >= 5)

# less than "<"
filter(iris, Sepal.Length < 5)

# equal & less than "<="
filter(iris, Sepal.Length <= 5)

```

`arrange()` : arrange the order of rows

```{r, eval=F}
# arrange in an ascending order
arrange(iris, Sepal.Length)

# arrange in an descending order
arrange(iris, desc(Sepal.Length))
```

### Column Manipulation {.unnumbered}

`select()`: select/remove column(s)

```{r, eval=F}
# select one column
select(iris, Sepal.Length)

# select multiple columns
select(iris, c(Sepal.Length, Sepal.Width))

# remove one column
select(iris, -Sepal.Length)

# remove multiple columns
select(iris, -c(Sepal.Length, Sepal.Width))

# select/remove multiple columns with a start rule
# starts_with("x")
select(iris, starts_with("Sepal"))
select(iris, -starts_with("Sepal"))

# select/remove multiple columns with an end rule
# ends_with("x")
select(iris, ends_with("Width"))
select(iris, -ends_with("Width"))
```

`mutate()`: add column(s)

```{r}
# add a new column
x <- 1:150
mutate(iris, x = x)
```

### Piping {.unnumbered}

`%>%` (pipe) allows sequential operations of multiple functions (shortcut: `Ctr + Shift + M`). The pipe passes the object to the following function as the first argument.

```{r}
# the following codes produce the same data frame
# apply functions separately
df_vir <- filter(iris, Species == "virginica")
df_vir_sl <- select(df_vir, Sepal.Length)
print(df_vir_sl)

# piping
iris %>% 
  filter(Species == "virginica") %>% 
  select(Sepal.Length)
```

### Reshape {.unnumbered}

`pivot_wider()` : reshape a data frame to a wide format

```{r}
iris_w <- iris %>% 
  mutate(id = rep(1:50, 3)) %>% # add an ID column
  select(id, Sepal.Length, Species) %>% 
  pivot_wider(id_cols = "id", # unique row ID based on
              values_from = "Sepal.Length", # values in each cell from
              names_from = "Species") # new column names from

print(iris_w)
```

`pivot_longer()`: reshape a data frame to a long format

```{r}
iris_l <- iris_w %>% 
  pivot_longer(cols = c("setosa",
                        "versicolor",
                        "virginica"), # columns with values to be reshaped
               names_to = "Species", # column IDs move to "Species"
               values_to = "Sepal.Length") # column values move to "Sepal.Length"

print(iris_l)
```

### Group Operation {.unnumbered}

`group_by()` & `summarize()`: group-by-group operation. `summarize()` does not retain individual rows.

```{r}
# grouping by "Species", then take means "Speal.Length" for each species
iris %>% 
  group_by(Species) %>% 
  summarize(mu_sl = mean(Sepal.Length))

# grouping by "Species", then take means & SD "Speal.Length" for each species
iris %>% 
  group_by(Species) %>% 
  summarize(mu_sl = mean(Sepal.Length),
            sd_sl = sd(Sepal.Length))
```

`group_by()` & `mutate()`: group-by-group operation. `mutate()` retains individual rows along with summary columns. Do not forget `ungroup()` to avoid errors in following operations.

```{r}
# grouping by "Species", then take means "Speal.Length" for each species
iris %>% 
  group_by(Species) %>% 
  mutate(mu_sl = mean(Sepal.Length)) %>% 
  ungroup()
```

### Join {.unnumbered}

`left_join()`: merge data frames based on column(s)

```{r}
# matching by a single column
## left join by "Species": one to one
df1 <- tibble(Species = c("A", "B", "C"),
              x = c(1, 2, 3))

df2 <- tibble(Species = c("A", "B", "C"),
              y = c(4, 5, 6))

left_join(x = df1,
          y = df2,
          by = "Species")

# matching by a single column
## left join by "Species": one to many
df3 <- tibble(Species = c("A", "A", "B", "C"),
              y = c(4, 5, 6, 7))

left_join(x = df1,
          y = df3,
          by = "Species")

# matching by a single column
## left join by "Species": one to missing
df4 <- tibble(Species = c("A", "A", "C"),
              y = c(4, 5, 7))

left_join(x = df1,
          y = df4,
          by = "Species")

# matching by multiple columns
## one to one
df5 <- tibble(Species = c("A", "B", "C"),
              x = c(1, 2, 3),
              z = c("cool", "awesome", "magical"))

left_join(x = df1,
          y = df5,
          by = c("Species", "x"))

# matching by multiple columns
## one to many
df6 <- tibble(Species = c("A", "A", "B", "C"),
              x = c(1, 1, 2, 3),
              z = c("cool", "cool", "awesome", "magical"))

left_join(x = df1,
          y = df6,
          by = c("Species", "x"))

# matching by multiple columns
## one to missing
df6 <- tibble(Species = c("A", "B", "C"),
              x = c(1, 2, 4),
              z = c("cool", "awesome", "magical"))

left_join(x = df1,
          y = df6,
          by = c("Species", "x"))
```

### Exercise {.unnumbered}

Complete the exercise [here](https://m-clark.github.io/data-processing-and-visualization/tidyverse.html#tidyverse-exercises).

## Visualization {.unnumbered}

`ggplot2::` offers a range of convenient functions for data visualization. The foundational function, `ggplot()`, provides the initial framework for adding supplementary layers using the `+` operator. In ggplot(), we define variables plotted on x- and y-axis through `aes()`. For example:

```{r, eval=F}
# without pipe
ggplot(data = iris,
       mapping = aes(x = Sepal.Length,
                     y = Sepal.Width)) +
  # additional layers...

# with pipe
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length,
                       y = Sepal.Width)) +
  # additional layers...  
```

Please note that `aes()` refers to columns in the data frame. Variables names that do not exist in the data frame cannot be used.

### Point {.unnumbered}

`geom_point()` : Add a point layer

```{r}
# basic plot
iris %>% 
  ggplot(aes(x = Sepal.Length,
             y = Sepal.Width)) +
  geom_point()

# change color by "Species" column
iris %>% 
  ggplot(aes(x = Sepal.Length,
             y = Sepal.Width,
             color = Species)) +
  geom_point()
```

### Line {.unnumbered}

`geom_line()` : Add a line layer

```{r}
# sample data
df0 <- tibble(x = rep(1:50, 3),
              y = x * 2)

# basic plot
df0 %>% 
  ggplot(aes(x = x,
             y = y)) +
  geom_line()
```

### Histogram {.unnumbered}

`geom_histogram()` : add a histogram layer

```{r}
# basic plot; bins = 30 by default
iris %>% 
  ggplot(aes(x = Sepal.Length)) +
  geom_histogram()

# change bin width
iris %>% 
  ggplot(aes(x = Sepal.Length)) +
  geom_histogram(binwidth = 0.5)

# change bin number
iris %>% 
  ggplot(aes(x = Sepal.Length)) +
  geom_histogram(bins = 50)
```

### Boxplot {.unnumbered}

`geom_boxplot()` : add a boxplot layer

```{r}
# basic plot
iris %>% 
  ggplot(aes(x = Species,
             y = Sepal.Length)) +
  geom_boxplot()

# change fill by "Species"
iris %>% 
  ggplot(aes(x = Species,
             y = Sepal.Length,
             fill = Species)) +
  geom_boxplot()

# change fill by "Species", but consistent color
iris %>% 
  ggplot(aes(x = Species,
             y = Sepal.Length,
             fill = Species)) +
  geom_boxplot(color = "darkgrey")
```

### Exercise {.unnumbered}

In maintenance
