# VISUALIZATION

## Overview

R has a number of functions that help visualize data. `graphics` provides R functions for base graphics ([list of functions](https://www.rdocumentation.org/packages/graphics/versions/3.6.2)). I will use `iris`, the built-in data set in R, to show how `graphics` functions work.

```{r}
iris
```

## `plot()`

When plotting data, what's you need is to specify the **formula**. For example, if you want to visualize the relationship between `x` and `y` (`y` on the vertical axis and `x` on the horizontal axis), the formula would be `y ~ x` (the left side of the formula will be on the vertical axis). In the `iris` data set, the following columns are available: `Sepal.Length` `Sepal.Width` `Petal.Length` `Petal.Width` `Species`. In the following example, the relationship between `Sepal.Length` and `Sepal.Width` is plotted:

```{r}

plot(Sepal.Length ~ Sepal.Width, data = iris)

```

The argument `data =` tells the function the data set from which the variables (`Sepal.Length` and `Sepal.Width`) are extracted. The above figure is the default setting, and you may customize it as necessary. Below is an example of how you may customize:

### Symbol

`pch` argument. Choose from `1` to `25` (google **r plot pch** for details)

```{r}

plot(Sepal.Length ~ Sepal.Width, data = iris,
     pch = 19)

```

### Symbol size

`cex` argument. `cex = 1` is the default value. `cex = 2` is as twice large as default value.

```{r}

plot(Sepal.Length ~ Sepal.Width, data = iris,
     pch = 19, cex = 2)

```

### Symbol color (border)

`col` argument (quote `"color name"` when specifying). Google **r color name** for color options.

```{r}

plot(Sepal.Length ~ Sepal.Width, data = iris,
     pch = 21, cex = 2, col = "gray")

```

### Symbol color (fill)

`bg` argument (quote `"color name"` when specifying). Available for a subset of symbol options (some symbols have pre-defined filled color).

```{r}

plot(Sepal.Length ~ Sepal.Width, data = iris,
     pch = 21, cex = 2, bg = "lightgray")

```

### Label

`ylab` or `xlab` arguments. Provide `"quoted text"`.

```{r}

plot(Sepal.Length ~ Sepal.Width, data = iris,
     pch = 21, cex = 2, bg = "lightgray",
     xlab = "Sepal width (cm)", ylab = "Sepal length (cm)")

```

### Axis

Delete axes with `axes = F` and re-draw with `box()` and `axis()` functions.

```{r}

plot(Sepal.Length ~ Sepal.Width, data = iris,
     pch = 21, cex = 2, bg = "lightgray",
     xlab = "Sepal width (cm)", ylab = "Sepal length (cm)",
     axes = F)
box(bty = "l") # L-shaped border lines
axis(1) # 1: draw x-axis
axis(2, las = 2) # 2: draw y-axis, las = 2: make axis lables horizontal

```

## `boxplot()`

`boxplot()` is used when the x-axis is factor-type data (by default, `plot()` will produce a boxplot when x-axis is a factor variable). In the `iris` dataset, the column `Species` is a factor variable. Compare `Sepal.Length` among species using `boxplot()`.

```{r}

boxplot(Sepal.Length ~ Species, data = iris)

```

You can customize as in `plot()`, but slighlty different.

### Box color

`col` argument.

```{r}

boxplot(Sepal.Length ~ Species, data = iris,
        col = "lightgray")

```

### Border color

`border` argument.

```{r}

boxplot(Sepal.Length ~ Species, data = iris,
        col = "lightgray", border = "grey48")

```

### Box width

`boxwex` argument.

```{r}

boxplot(Sepal.Length ~ Species, data = iris,
        col = "lightgray", border = "grey48",
        boxwex = 0.4 )

```

### Axis

Delete axes with `axes = F` and re-draw with `box()` and `axis()` functions.

```{r}

boxplot(Sepal.Length ~ Species, data = iris,
        col = "lightgray", border = "grey48",
        boxwex = 0.4, ylab = "Sepal length (cm)",
        axes = F)
box(bty = "l")
axis(1, at = c(1, 2, 3), labels = c("Setosa", "Versicolor", "Virginica") )
axis(2, las = 2)

```

## Exercise

Download a <a href="file/report02_temp.R" download>template</a> (R script file).

### `plot()` function

a.  Plot the relationship between `Sepal.Width` and `Petal.Width`
b.  Turn symbol color (border) into red.
c.  Turn symbol color (fill) into red (set `pch = 21`).
d.  Make symbol size larger.
e.  Make L-shaped plot border (delete border lines on upper and right sides)

### `boxplot()` function

a.  Plot the relationship between `Petal.Width` and `Species`.
b.  Turn box color (border) into `blue`.
c.  Make box width narrower.
