# Appendix: Base Plot {.unnumbered}

## Overview

R offers a variety of functions that aid in visualizing data. The `graphics` package in R provides a set of functions for basic graphics ([list of functions](https://www.rdocumentation.org/packages/graphics/versions/3.6.2)). To demonstrate the functionality of these `graphics` functions, I will utilize the built-in `iris` dataset in R.

```{r}
head(iris)
```

## Plot

When creating a plot, you typically need to specify the **formula** to define the relationship between variables. For instance, if you wish to visualize the association between `x` and `y` (with `y` on the vertical axis and `x` on the horizontal axis), the formula would be `y ~ x` (where the left side of the formula represents the vertical axis). In the `iris` dataset, you have access to the following columns: `Sepal.Length`, `Sepal.Width`, `Petal.Length`, `Petal.Width`, and `Species`. In the subsequent example, we will plot the relationship between `Sepal.Length` and `Sepal.Width`:

```{r}

plot(Sepal.Length ~ Sepal.Width, data = iris)

```

The `data =` argument informs the function about the dataset from which the variables (`Sepal.Length` and `Sepal.Width`) should be extracted.

### Symbol

`pch` argument. Choose from `1` to `25` (google **r plot pch** for details)

```{r}

plot(Sepal.Length ~ Sepal.Width, data = iris,
     pch = 19)

```

### Symbol size

`cex` argument. `cex = 1` is the default value. `cex = 2` is as twice large as default value.

```{r}

plot(Sepal.Length ~ Sepal.Width, data = iris,
     pch = 19, cex = 2)

```

### Symbol color (border)

`col` argument (quote `"color name"` when specifying). Google **r color name** for color options.

```{r}

plot(Sepal.Length ~ Sepal.Width, data = iris,
     pch = 21, cex = 2, col = "gray")

```

### Symbol color (fill)

`bg` argument (quote `"color name"` when specifying). Available for a subset of symbol options (some symbols have pre-defined filled color).

```{r}

plot(Sepal.Length ~ Sepal.Width, data = iris,
     pch = 21, cex = 2, bg = "lightgray")

```

### Label

`ylab` or `xlab` arguments. Provide `"quoted text"`.

```{r}

plot(Sepal.Length ~ Sepal.Width, data = iris,
     pch = 21, cex = 2, bg = "lightgray",
     xlab = "Sepal width (cm)", ylab = "Sepal length (cm)")

```

### Axis

Delete axes with `axes = F` and re-draw with `box()` and `axis()` functions.

```{r}

plot(Sepal.Length ~ Sepal.Width, data = iris,
     pch = 21, cex = 2, bg = "lightgray",
     xlab = "Sepal width (cm)", ylab = "Sepal length (cm)",
     axes = F)
box(bty = "l") # L-shaped border lines
axis(1) # 1: draw x-axis
axis(2, las = 2) # 2: draw y-axis, las = 2: make axis lables horizontal

```

## Boxplot

`boxplot()` is used when the x-axis is factor-type data (by default, `plot()` will produce a boxplot when x-axis is a factor variable). In the `iris` dataset, the column `Species` is a factor variable. Compare `Sepal.Length` among species using `boxplot()`.

```{r}

boxplot(Sepal.Length ~ Species, data = iris)

```

You can customize as in `plot()`, but slighlty different.

### Box color

`col` argument.

```{r}

boxplot(Sepal.Length ~ Species, data = iris,
        col = "lightgray")

```

### Border color

`border` argument.

```{r}

boxplot(Sepal.Length ~ Species, data = iris,
        col = "lightgray", border = "grey48")

```

### Box width

`boxwex` argument.

```{r}

boxplot(Sepal.Length ~ Species, data = iris,
        col = "lightgray", border = "grey48",
        boxwex = 0.4 )

```

### Axis

Delete axes with `axes = F` and re-draw with `box()` and `axis()` functions.

```{r}

boxplot(Sepal.Length ~ Species, data = iris,
        col = "lightgray", border = "grey48",
        boxwex = 0.4, ylab = "Sepal length (cm)",
        axes = F)
box(bty = "l")
axis(1, at = c(1, 2, 3), labels = c("Setosa", "Versicolor", "Virginica") )
axis(2, las = 2)

```
