fish density in a lake with a total area of 225 square meters (15 meters by 15 meters). To measure fish density, we choose a specific area of 1 square meter for sampling. However, there might be variation in fish density among different sampled plots within the lake. To account for this variability, we repeat the *sampling* process 10 times. Each repetition involves randomly selecting a 1 square meter area and counting the number of fish within that area. By conducting these repeated samples, we can calculate the average fish density across the lake. Figure \@ref(fig:fish-plot) shows a graphical representation of this

```{r fish-plot, fig.cap="Lake fish density example. Points are fish, and white boxes are sampling sites with 1 sq-m"}

knitr::include_graphics(here::here("image/figure_lake.pdf"))
```

Through this sampling, we got the following dataset:

```{r df_count, echo = FALSE}

id <- 1:nrow(df_plot)
plot100 <- c(plot10, sample(id[!(id %in% plot10)], size = 90))

df_count_full <- df_fish_sub %>% 
  distinct(plot, count) %>% 
  arrange(plot) %>% 
  slice(plot100)

df_count <- df_count_full %>% 
  filter(plot %in% plot10) %>% 
  mutate(plot = row_number())
```

```{r df_count_print, echo = TRUE}
df_count
```

Based on this dataset, we may conclude the average fish density in this lake is `r df_count %>% pull(count) %>% mean() %>% round(2)`. However, how confident are you in this conclusion? The lake area is $15 \times 15 = 225~\text{m}^2$ but we sampled only $1~\text{m}^2 \times 10 = 10~\text{m}^2$! Indeed, we missed a lot of fish in the lake (Figure \@ref(fig:fish-plot-2)).

```{r fish-plot-2, fig.cap="True fish distribution in the lake"}
g_lake + 
  geom_point(data = df_fish,
             size = 0.5,
             color = grey(0, 0.2))
```

In ecological research (and almost all research fields), we are rarely able to obtain the perfect dataset with no uncertainty -- the data are partial information of what we want to know (true fish density in the entire lake = `r round(nrow(df_fish)/(15^2), 2)`). This is the reason why we refer to a data point as a **sample**, and the average fish density in a plot (`r df_count %>% pull(count) %>% mean() %>% round(2)`) is the **estimate** of the true fish density -- it is the inference from the incomplete information. Since the average fish density is the estimate, it does not perfectly match the true density -- how do we account for this uncertainty?
